An enhancer code compendium of early human neural and neural crest development

### 2025-04-18 manuscript.tex
Move parts of figure 3 to supplementary figure 4

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. 
    \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIP-seq peaks (general; bottom) and those regions that do not overlap (specific; top). 
    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 2,130) and specific (n = 4,529) regions, for organoid.
    \textbf{k}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions. * indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region. 
    \textbf{l}, de novo motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
 }
\end{figure}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) 
the importance of TEAD instances correlates with \textit{TEAD3/4} expression, 
GRHL with \textit{GRHL1/1} 
and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
and moreover a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). 
Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). 
We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. 
The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), 
ZIC with \textit{ZIC1} to \textit{ZIC5} and 
the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. 
After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) 
for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. 
Instances of a FOX patterns and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). 
The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

@@

Recently, a genome wide association study (GWAS) identified 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
we asked if, using $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ 
and calculated the difference in prediction score between both alleles (delta prediction score). 
At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S6}). 
As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
(\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme. 
This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimer (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neural progenitors are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}). 
The importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}); 
instead we find topics where a combination of EBOX and EBF instances are important 
(organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
(\textbf{Fig. 6d} and Supplementary Fig. S7) 
and topics where a combination of EBF and ONECUT instances are important 
(organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), 
the latter correlating with the expression of \textit{ONECUT1-3}. 
Additionally, in the embryo we found topics where instances of NKX are important. 
The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
Finally, in the organoids we could identify an additional topic with instances of GATA
(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate. That you don't see it in the organoid might indicate that part of this patterning is not present. https://www.nature.com/articles/19315
%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning? Are all the same types of neurons there or might it be related to a missing cell type?

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in 
respectively the neural tube organoids and the head of a four weeks post conception human embryo with 
almost one-to-one correspondence based on TFBS importances and TF expression. 
Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neural progenitors 
in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
logistic regression models to classify genomic regions as belonging to the topic representing each state or not,
making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). We used regions of the topic of interest as foreground and regions of the other topics of the same cell type-class as background.
The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average auROC of 0.77 
and average auPR of 0.39. 
Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S8). \par

@@

with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 9}). 

@@

In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. 
The effect a TF has on target gene expression depends on its effector domain and interactions with
co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. 
Large scales studies, in a variety of cell states, are now being performed to map TFs to co-factors and downstream effects on gene expression\cite{
  tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
sotoCompendiumHumanTranscription2022, delrossoLargescaleMappingMutagenesis2023,
mukundHighthroughputFunctionalCharacterization2023, bellComparativeCofactorScreens2024,
tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. 
These studies may especially be important to infer repressive roles of TFs on gene expression, 
an effect that is currently difficult to study with state of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
dewinterModellingDesignTranscriptional2025}. 
In particular for the neural tube, it is known that repressive interactions in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. 
In this study we could only predict a strong repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
(\textbf{Supplementary Fig. S10}).\par

@@

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_4.png}
  \caption{\textbf{Linking motif to TF in anterior-posterior neural progenitors}.\\
    \textbf{a-e} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
    that is annotated to: Homeodomain or Homeodomain; Paired box or Paired box (\textbf{a-e}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
  }
\end{figure}

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_5.png}
  \caption{
    \textbf{Linking motif to TF in neural crest}.\\
    \textbf{a-j} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
    that is annotated to:
    Forkhead (\textbf{a}), 
    HMG/SOX (\textbf{b}), 
    TEA (\textbf{c}), 
    AP-2 (\textbf{d}), 
    C2H2 ZF; Homeodomain (\textbf{e}), 
    C2H2 ZF (\textbf{f}), 
    Nuclear receptor (\textbf{g}), 
    Grainyhead (\textbf{h}), 
    bHLH (\textbf{i}), 
    RFX (\textbf{j}).
    Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
  }
\end{figure}
\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_6.png}
  \caption{
    \textbf{Delta prediction score across topics and SNPs}\\
    \textbf{a}, heatmap showing delta prediction score across topics using $DeepNeuralTube_O$ (left) and
    $DeepNeuralTube_E$ (right) for SNPs with a minimal absolute delta prediction score of 0.2.
  }
\end{figure}

\afterpage{\clearpage}
\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_7.png}
  \caption{
    \textbf{Linking motif to TF in neurons}.\\
    (Continues next page):
 }
\end{figure}
\afterpage{\clearpage}

>>>>

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
}
\end{figure}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Supplementary Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) 
the importance of TEAD instances correlates with \textit{TEAD3/4} expression, 
GRHL with \textit{GRHL1/1} 
and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
and moreover a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). 
Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). 
We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. 
The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), 
ZIC with \textit{ZIC1} to \textit{ZIC5} and 
the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. 
After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) 
for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S6).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. 
Instances of a FOX patterns and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). 
The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

@@

Recently, a genome wide association study (GWAS) identified 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
we asked if, using $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ 
and calculated the difference in prediction score between both alleles (delta prediction score). 
At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S7}). 
As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
(\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme. 
This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimer (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neural progenitors are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}). 
The importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}); 
instead we find topics where a combination of EBOX and EBF instances are important 
(organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
(\textbf{Fig. 6d} and Supplementary Fig. S8) 
and topics where a combination of EBF and ONECUT instances are important 
(organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), 
the latter correlating with the expression of \textit{ONECUT1-3}. 
Additionally, in the embryo we found topics where instances of NKX are important. 
The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
Finally, in the organoids we could identify an additional topic with instances of GATA
(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S8).\par
%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate. That you don't see it in the organoid might indicate that part of this patterning is not present. https://www.nature.com/articles/19315
%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning? Are all the same types of neurons there or might it be related to a missing cell type?

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in 
respectively the neural tube organoids and the head of a four weeks post conception human embryo with 
almost one-to-one correspondence based on TFBS importances and TF expression. 
Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neural progenitors 
in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
logistic regression models to classify genomic regions as belonging to the topic representing each state or not,
making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). We used regions of the topic of interest as foreground and regions of the other topics of the same cell type-class as background.
The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average auROC of 0.77 
and average auPR of 0.39. 
Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S9). \par

@@

with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 10}). 

@@

In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. 
The effect a TF has on target gene expression depends on its effector domain and interactions with
co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. 
Large scales studies, in a variety of cell states, are now being performed to map TFs to co-factors and downstream effects on gene expression\cite{
  tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
sotoCompendiumHumanTranscription2022, delrossoLargescaleMappingMutagenesis2023,
mukundHighthroughputFunctionalCharacterization2023, bellComparativeCofactorScreens2024,
tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. 
These studies may especially be important to infer repressive roles of TFs on gene expression, 
an effect that is currently difficult to study with state of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
dewinterModellingDesignTranscriptional2025}. 
In particular for the neural tube, it is known that repressive interactions in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. 
In this study we could only predict a strong repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
(\textbf{Supplementary Fig. S11}).\par

@@

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_4.png}
  \caption{
    \textbf{a}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. 
    \textbf{b}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIP-seq peaks (general; bottom) and those regions that do not overlap (specific; top). 
    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 2,130) and specific (n = 4,529) regions, for organoid.
    \textbf{c}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions. * indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region. 
  }
\end{figure}

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_5.png}
  \caption{\textbf{Linking motif to TF in anterior-posterior neural progenitors}.\\
    \textbf{a-e} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
    that is annotated to: Homeodomain or Homeodomain; Paired box or Paired box (\textbf{a-e}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
  }
\end{figure}

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_6.png}
  \caption{
    \textbf{Linking motif to TF in neural crest}.\\
    \textbf{a-j} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
    that is annotated to:
    Forkhead (\textbf{a}), 
    HMG/SOX (\textbf{b}), 
    TEA (\textbf{c}), 
    AP-2 (\textbf{d}), 
    C2H2 ZF; Homeodomain (\textbf{e}), 
    C2H2 ZF (\textbf{f}), 
    Nuclear receptor (\textbf{g}), 
    Grainyhead (\textbf{h}), 
    bHLH (\textbf{i}), 
    RFX (\textbf{j}).
    Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
  }
\end{figure}
\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_7.png}
  \caption{
    \textbf{Delta prediction score across topics and SNPs}\\
    \textbf{a}, heatmap showing delta prediction score across topics using $DeepNeuralTube_O$ (left) and
    $DeepNeuralTube_E$ (right) for SNPs with a minimal absolute delta prediction score of 0.2.
  }
\end{figure}

\afterpage{\clearpage}
\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_8.png}
  \caption{
    \textbf{Linking motif to TF in neurons}.\\
    (Continues next page):
 }
\end{figure}
\afterpage{\clearpage}
